### 2020-06-21 article.tex
Figures
start to work on the paper (up to subsection 2.2.1 included)

% Optional PDF information
\ifpdf
\hypersetup{
  pdftitle={KSPHPDDM and PCHPDDM: Two Classes of Adaptive Krylov and Overlapping Schwarz Methods in PETSc},
  pdfauthor={P. Jolivet and S. Zampini}
}
\fi

@@

% REQUIRED
\begin{abstract}
  Contemporary applications in computational science and engineering often
    require the solution of linear systems which may be of different sizes,
    shapes, and structures. The goal of this paper is to explain how two
    libraries, PETSc and HPDDM, have been interfaced together in order to offer
    to end-users adaptive Krylov and overlapping Schwarz methods.  The
    flexibility of the implementation %, which for example may deal with various matrix formats, 
    is showcased and explained.  Diverse examples are
    displayed, ranging from fracture mechanics to linear stability analysis, and
    the sets of options are thoroughly documented, so that they may be readily
    tried on other applications and reproduce results.
\end{abstract}

@@

\section{Introduction}
Implementing and maintaining robust and scalable applications is highly relevant
in the field of computational science and engineering. Some tools exist to help
end-users achieve this goal. One the one hand, PETSc \cite{balay1997cient}, the
Portable and Extensible Toolkit for Scientific computation, is a
well-established software from the community. It may be used to efficiently
discretize partial differential equations, or solve algebraic linear or
nonlinear systems of equations. It comes with many built-in features such as
advanced preconditioning or tailored matrix formats. Also, it can % be used to compile and 
interact with third-party libraries, such as
\emph{hypre}~\cite{falgout2002hypre} for linear solvers, TetGen~\cite{si2013tet}
for mesh generation, and so on. These appealing features convinced computational
scientists to use PETSc as one of the discretization and/or algebraic backend in
many different higher-level projects, such as finite element or finite volume
packages. On the other hand, HPDDM is a much smaller project purely focused on
domain decomposition preconditioners~\cite{jolivet2013scalable} % , a small subclass of preconditioners available in PETSc as well, 
and advanced iterative
methods~\cite{jolivet2016block}. Such methods deal efficiently with linear
systems when multiple right-hand sides are available simultaneously, and are
then referred to as block iterative methods, or when there is a recurrence of
varying coefficient matrices and right-hand sides, and are then referred to as
recycling iterative methods.
% Htool ?
In this paper, the integration of both aspects of HPDDM, its advanced iterative
methods and its domain decomposition preconditioners, into PETSc is explained.
The design principles of the interface are showcased, with % alongside
minimalist and easy-to-run examples. The paper is divided in two main parts:
\cref{sec:KSPHPDDM} which introduces \pk{KSPHPDDM}, and \cref{sec:PCHPDDM} which
introduces \pk{PCHPDDM}. Eventually, concluding remarks may be found
\cref{sec:conclusion}. All PETSc keywords and options are typeset in a \texttt{typewriter} front, these
are public and documented at \url{https://www.mcs.anl.gov/petsc/petsc-master/docs}.
\section{KSPHPDDM}\label{sec:KSPHPDDM}
  \subsection{Related work}
Krylov subspace methods are widely used in numerical linear algebra for solving
systems of equations. Some of their strengths are memory efficiency and the fact
that they are mostly format agnostic. Indeed, they mainly rely on matrix--vector
operations such as multiplications or transpose multiplications. For
challenging high-dimensional systems, finding appropriate preconditioners is
mandatory to make these methods robust. %, but this goes beyond the scope of this section.
In PETSc, Krylov methods belong to the \pk{KSP} base class. Some of the
most used types are \pk{KSPGMRES}, \pk{KSPCG}, or \pk{KSPBCGS}, which respectively implement
the generalized minimal residual method~\cite{saad1986gmres}~(GMRES), conjugate
gradient~\cite{hestenes1952methods}~(CG), or biconjugate gradient stabilized method~\cite{van1992bi}~(BCGS). % Some of these methods are also
% available in other linear algebra backends such as~\emph{hypre} or DUNE~\cite{bastian2010generic}.

Because these methods may need long-term recurrences to mitigate round-off
errors during the generation of Krylov subspaces, it is common to introduce a
restart parameter in order to control their memory consumption and the amount of
global communication, e.g., when orthonormalizing a candidate basis vector. Such
restarts may hinder convergence of iterative methods, and even introduce
convergence plateaus. Recycling techniques have been introduced to attenuate
these effects. In PETSc, the Loose GMRES~\cite{baker2005technique}~(LGMRES) and the
Deflated GMRES~\cite{erhel1996restarted,wakam2013158}~(DGMRES), implemented as
\pk{KSPLGMRES} and \pk{KSPDGMRES}, are available. Neither handle variable
preconditioning, and the latter does not support complex arithmetic. Also note
that when solving a sequence of linear systems $A_i x_i = b_i, i = 1, 2,
\ldots$, it is not trivial to extend the theory of these methods to perform
recycling throughout both the restarts and the different systems. The
generalized conjugate residual method with inner orthogonalization and deflated
restarting~\cite{parks2006recycling}~(GCRODR) does not suffer from this
limitation.

% Recycling put aside, 
Another important aspect of Krylov methods is their ability
to deal with multiple right-hand sides simultaneously. Block Krylov methods~\cite{gutknecht2006block} are
designed for solving linear systems $A X = B$, where
$X$ and $B$ are tall-and-skinny matrices with $k \geq 1$ columns. These methods are
less frequently implemented in general purpose libraries because they require
somehow more involved kernels such as matrix--matrix multiplication, instead of
matrix--vector. Still, systems with multiple right-hand sides are ubiquitous,
e.g., in tomography~\cite{tournier2016micro}, data
analytics~\cite{KALANTZIS2018136}, eigensolvers~\cite{knyazev2001toward}, geophysics~\cite{calandra2012flexible}, quantum chromodynamics~\cite{sakurai2010application}, etc. % Of course, it is possible to cast a system with multiple right-hand sides into a sequence of systems with the same coefficient matrix but changing right-hand sides $A x_i = b_i, i = 1,2,\ldots,k$. However, 
Block Krylov methods, while having higher arithmetic
intensities, generate larger subspaces and typically converge in fewer iterates.
There is none implemented in PETSc. It is nevertheless possible to solve
systems with multiple right-hand sides using a specialized type of matrices,
but with a very limited choice of preconditioners, cf.~\cref{sec:matkaij}.
In the context of optimization, it is also
frequent to solve a sequence of linear systems $A_i X_i = B_i, i = 1, 2,
\ldots$

Trilinos~\cite{heroux2005overview} is another well-known software for scientific
computing. Through its Belos package~\cite{bavier2012amesos2}, it provides most
of the advanced iterative methods aforementioned. Although having PETSc and
Trilinos interoperate is possible, there is currently no \pk{KSP} interface to Belos. % This limits its composability inside complete PETSc applications.
Furthermore,
as displayed in~\cref{sec:interface-ksp}, some Krylov methods implemented in
HPDDM are not available in Belos. In the following section, the details of the
interface between HPDDM Krylov methods and PETSc matrix and preconditioner
classes will be explained. It is assumed that the right-hand sides, and
consequently the solutions, are always dense vectors or matrices.
  \subsection{Interfacing HPDDM Krylov methods in PETSc\label{sec:interface-ksp}}\
\pk{KSPHPDDM}, the interface between HPDDM Krylov methods and PETSc, is automatically
registered since PETSc version 3.12 when configuring PETSc with the extra
flag \pk{-{}-download-hpddm}. It is then possible to select the corresponding \pk{KSP}
using the command line option \pk{-ksp\_type hpddm}, or the routine \pk{KSPSetType}(ksp,
\pk{KSPHPDDM}).% and its Fortran equivalent.
When \pk{KSPHPDDM} is being used, it is then possible to select between the following
Krylov methods:
    \item pseudo-block GMRES or flexible GMRES~\cite{saad1993flexible};
    \item pseudo-block CG or flexible CG~\cite{notay2000flexible};
    \item pseudo-block GCRODR or flexible GCRODR~\cite{carvalho2011flexible};
    \item block GMRES or flexible GMRES~\cite{calandra2012flexible};
    \item block CG~\cite{o1980block};
    \item breakdown-free block CG~\cite{ji2017breakdown};
    \item block GCRODR or flexible GCRODR.
In pseudo-block variants, multiple similar operations like matrix--vector
products are fused to achieve higher arithmetic intensity, or to decrease the
number of global synchronizations, e.g., reductions for scalar products. But they are mathematically equivalent to
their ``standard'' counterpart. These methods may be selected using the 
options \pk{-ksp\_hpddm\_type} \pk{(gmres$|$cg$|$gcrodr$|$bgmres$|$bcg$|$bfbcg$|$bgcrodr$|$preonly)} and
\pk{-ksp\_hpddm\_variant} \pk{(left$|$right$|$flexible)} to select whether preconditioning
is applied on the left, on the right, or variable. It is also possible to set
this option through the more common PETSc option \pk{-ksp\_pc\_side (left$|$right)}. 
Note that the value of the latter option is useless for BCG or BFBCG, and HPDDM CG only handles left and
variable preconditioning. According to the standard PETSc notation, ksp($A$, $P_B$)
defines a \pk{KSP} used to solve a system $Ax=b$ using a preconditioner $P_B$ built
using an operator $B$ which is usually $A$ but not necessarily. If the
type of ksp is \pk{KSPHPDDM}, the solution phase, invoked by \pk{KSPSolve}(ksp, $b$, $x$), will be carried out by HPDDM. To do
so, HPDDM will repeatedly call the following PETSc routines:
    \item \pk{MatMult}($A$, $x$, $y$) for $y = Ax$;
    \item \pk{MatMatMult}($A$, $X$, {\scriptsize \pk{MAT\_REUSE\_MATRIX}}, {\scriptsize
        \pk{PETSC\_DEFAULT}}, $Y$)\footnote{or \pk{MatProductNumeric}($Y$) with PETSc 3.14.0 and above} for $Y =
        AX$;
    \item \pk{PCApply}($P_B$, $y$, $x$) for $x = P_By$.
Currently, PETSc does not have a general routine for applying $P_B$ to a
tall-and-skinny dense matrix $Y$. In HPDDM, a specialized routine has been implemented
to compute $X = P_BY$ if the preconditioner is of one of the following types:
    \item exact $LU$ or Cholesky factorization, i.e., $P_B = B^{-1}$, \pk{PCLU} or
        \pk{PCCHOLESKY};
    \item incomplete $LU$ or Cholesky factorization, \pk{PCILU} or \pk{PCICC};
    \item block Jacobi with % a single subdomain per process and
        exact or incomplete subdomain solvers, \pk{PCBJACOBI};
    \item overlapping Schwarz methods with % a single subdomain per process and
        exact or incomplete subdomain solvers, \pk{PCASM}.
Indeed, in these cases, it is possible to get access to a factored matrix $F$
and call \pk{MatMatSolve}($F$, $\tilde{Y}$, $\tilde{X}$). % With the two first sets of preconditioners, $\tilde{Y} = Y$ and $\tilde{X} = X$. With block Jacobi, $\tilde{Y}$, resp. $\tilde{X}$, represents only the part of the matrix $Y$, resp. $X$, local to a given process. With overlapping Schwarz methods, there are furthermore overlapping ``ghost'' unknowns, interested readers are referred to monographs on practical aspects of domain decomposition methods~\cite{smith2004domain,dolean2014ddm}.
For all other types of preconditioners, HPDDM will loop over the columns of $Y$
and apply $P_B$ successively to get all columns of $X$.

The rest of the operations are then performed directly inside HPDDM. For
example, in all Krylov methods using the Arnoldi process, it is mandatory to orthonormalize
candidate basis vectors. % It would be impractical to try to reuse PETSc routines like \pk{KSPGMRESModifiedGramSchmidtOrthogonalization} for two reasons: 1) it does not handle the orthonormalization of multiple vectors at once which is needed in pseudo-block methods 2) HPDDM may be used without PETSc as a stand-alone library. To give some more technical details to curious readers, in HPDDM,
$QR$ factorizations of tall-and-skinny dense matrices, which are needed by
block Krylov methods and for recycling Krylov methods are computed using
the CholQR algorithm~\cite{stathopoulos2002block}, or the block modified or classical
Gram--Schmidt method. This may be adjusted with the option \pk{-ksp\_hpddm\_qr (cholqr$|$mgs$|$cgs)}.
For Hessenberg matrices generated by the Arnoldi
process, it is common to update their $QR$ decomposition using Givens
rotations~\cite{saad2003iterative}. For block Hessenberg matrices, Householder reflectors are
used instead~\cite{gutknecht2008updating}.

Monitoring convergence with the \pk{KSPMonitor} and \pk{KSPSetConvergenceTest} callbacks is fully supported.

    \subsubsection{User-defined deflation space} Recycling Krylov methods try to
extract information at the end of each cycle or when convergence is reached by
solving small standard or generalized dense eigenproblems. Then, during a
subsequent solve, this information is reused appropriately~\cite{saad2000deflated,stathopoulos2002block,soodhalter2014krylov}. Two new routines were
added in PETSc to either get the information generated by such recycling
methods, or set the information in case users know what is a good initial
deflation subspace. These routines are respectively
\pk{KSPHPDDMGetDeflationSpace} and \pk{KSPHPDDMSetDeflationSpace}. The subspaces
are once again stored as distributed
dense tall-and-skinny matrices.

@@

@@

@@

>>>>

% Optional PDF information
\ifpdf
\hypersetup{
  pdftitle={KSPHPDDM and PCHPDDM: extending PETSc with robust Overlapping Schwarz Methods and advanced Krylov methods},
  pdfauthor={P. Jolivet and S. Zampini}
}
\fi

@@

% REQUIRED
\begin{abstract}
  Contemporary applications in computational science and engineering often
    require the solution of linear systems which may be of different sizes,
    shapes, and structures. The goal of this paper is to explain how two
    libraries, PETSc and HPDDM, have been interfaced together in order to offer
    end-users robust overlapping Schwarz methods and advanced Krylov methods featurin
    recycling and block solution techniques.
    The flexibility of the implementation %, which for example may deal with various matrix formats,
    is showcased and explained with minimalist, easy-to-run and reproducible examples
    ranging from {\color{red}fracture mechanics to linear stability analysis}
    to ease the integration of these algorithms into more advanced frameworks.
\end{abstract}

@@

\section{Introduction}
Implementing and maintaining robust and scalable applications is highly relevant
in the field of computational science and engineering. The efficient production of
such applications requires a rich ecosystem of state-of-the-art reusable libraries
\cite{xsdk}. In this paper, we focus on the interoperability of two such libraries.

One the one hand, PETSc \cite{balay1997cient}, the
Portable and Extensible Toolkit for Scientific computation, is a
well-established, actively developed software from the community. It may be used to efficiently
discretize partial differential equations and solve algebraic linear or
nonlinear systems of time-dependent equations. Among its strenghts, it offers many
advanced features regarding preconditioning and tailored matrix formats, and it can
interact with third-party libraries, such as
\emph{hypre}~\cite{falgout2002hypre} for linear solvers, TetGen~\cite{si2013tet}
for mesh generation, p4est~\cite{BursteddeWilcoxGhattas11} for adaptive mesh
refinement, to cite a few. The extensibility and robustness of the framework
convinced computational scientists to use PETSc as one of the discretization and/or algebraic backend in
many different higher-level projects {\color{red} ADD CIT TO DEAL.II, FIREDRAKE, MOOSE, HIPPYLIB, MFEM}.

On the other hand, the High-Performance unified framework for Domain Decomposition Methods (HPDDM)
is a much smaller project focusing on robust and scalable domain decomposition
preconditioners~\cite{jolivet2013scalable} and advanced iterative
methods~\cite{jolivet2016block} which can efficiently deal with linear
systems with multiple right-hand sides (also known and as block iterative methods)
and when there is a recurrence of varying coefficient matrices and right-hand
sides (recycling iterative methods). For a introduction to domain decomposition methods, see
~\cite{smith2004domain,dolean2014ddm,olofbook,tarekmathewbook} {\color{red} FIX CITATIONS}.

% Htool ?
Here we describe the integration of the entire HPDDM suite, its advanced iterative
methods and its domain decomposition preconditioners, into PETSc.
We showcase the design principles of the interface considering % alongside
minimalist and easy-to-run examples, equipped with the necessary command line options to
fullfill reproducibility requirements {\color{red} FIND GOOD CIT, ADD 1 more sentences?,
https://www.acm.org/publications/policies/artifact-review-badging} and ease the
integration of these algorithms into more advanced frameworks.

The paper is divided in two main parts {\color{red} COMPLETE WITH MORE DETAILS WHEN PAPER IS COMPLETED}:
\cref{sec:KSPHPDDM} introduces \pk{KSPHPDDM}, provides applications within an eigensolver
framework and discuss the readiness of the PETSc library in delivering highly efficient block Krylov methods.
\cref{sec:PCHPDDM} introduces \pk{PCHPDDM}. Eventually, concluding remarks may be found
\cref{sec:conclusion}. All PETSc keywords and options are typeset in a \texttt{typewriter} front, these
are public and documented at \url{https://www.mcs.anl.gov/petsc/petsc-master/docs}.

\section{KSPHPDDM}\label{sec:KSPHPDDM}
  \subsection{Related work}
Krylov subspace methods are widely used in numerical linear algebra for solving
linear systems of equations because of their memory efficiency~\cite{templateBook}.
They mainly rely on matrix--vector
operations such as multiplications or transpose multiplications and,
for robust convergence when dealing with challenging high-dimensional systems,
on the application of appropriate preconditioners.
PETSc (as of version 3.13.2) already offers 44 different types of Krylov methods within the \pk{KSP} base class. Some of the
most used types are \pk{KSPGMRES}, \pk{KSPCG}, or \pk{KSPBCGS}, which respectively implement
the generalized minimal residual method~\cite{saad1986gmres}~(GMRES), conjugate
gradient~\cite{hestenes1952methods}~(CG), or biconjugate gradient stabilized method~\cite{van1992bi}~(BCGS). % Some of these methods are also
% available in other linear algebra backends such as~\emph{hypre} or DUNE~\cite{bastian2010generic}.

Because Krylov methods may need long-term recurrences to mitigate round-off
errors during the generation of subspaces, it is common to introduce a
restart parameter in order to control their memory consumption and the volume of
global communications needed when orthonormalizing a candidate basis vector. Such
restarts may hinder convergence of iterative methods, and even introduce
convergence plateaus. Recycling techniques have been introduced to attenuate
these effects. In PETSc, the Loose GMRES~\cite{baker2005technique}~(LGMRES) and the
Deflated GMRES~\cite{erhel1996restarted,wakam2013158}~(DGMRES), implemented as
\pk{KSPLGMRES} and \pk{KSPDGMRES}, are available. However, neither handle variable
preconditioning, and the DGMRES does not support complex arithmetic. {\color{red} Also note
that when solving a sequence of linear systems $A_i x_i = b_i, i = 1, 2,
\ldots$, it is not trivial to extend the theory of these methods to perform
recycling throughout both the restarts and the different systems. The
generalized conjugate residual method with inner orthogonalization and deflated
restarting~\cite{parks2006recycling}~(GCRODR) does not suffer from this
limitation.}

% Recycling put aside,
Another important aspect of Krylov methods is their ability
to deal with multiple right-hand sides simultaneously.
Block Krylov methods~\cite{gutknecht2006block} are
designed for solving linear systems $A X = B$, where
$X$ and $B$ are tall-and-skinny matrices with $k \geq 1$ columns.
Such methods, while having higher arithmetic
intensities, generate larger subspaces and typically converge in fewer iterates.

These methods are not currently offered in PETSc and they are
less frequently implemented in general purpose libraries because they require
somehow more involved kernels such as matrix--matrix multiplication, instead of
matrix--vector, and may involve different inner-products realizations~\cite{Frommer}.
Still, systems with multiple right-hand sides are ubiquitous,
e.g., in tomography~\cite{tournier2016micro}, data
analytics~\cite{KALANTZIS2018136}, eigensolvers~\cite{knyazev2001toward}, geophysics~\cite{calandra2012flexible}, quantum chromodynamics~\cite{sakurai2010application}, and optimization with time-dependent partial differential equations as constraints.
% Of course, it is possible to cast a system with multiple right-hand sides into a sequence of systems with the same coefficient matrix but changing right-hand sides $A x_i = b_i, i = 1,2,\ldots,k$. However,
{\color{red} REMOVE? It is nevertheless possible to solve
systems with multiple right-hand sides using a specialized type of matrices,
but with a very limited choice of preconditioners, cf.~\cref{sec:matkaij}.}

Trilinos~\cite{heroux2005overview} is another well-known software for scientific
computing and it provides iterative methods through its Belos package~\cite{bavier2012amesos2},
Although having PETSc and
Trilinos interoperate is possible, there is currently no \pk{KSP} interface to Belos. % This limits its composability inside complete PETSc applications.
Furthermore, some of the Krylov methods implemented in
HPDDM and discussed in ~\cref{sec:interface-ksp}, are not available in Belos.

  \subsection{Interfacing HPDDM Krylov methods in PETSc\label{sec:interface-ksp}}\
In this section we provide details of the
interface between HPDDM Krylov methods and the various PETSc
classes. It is assumed that the right-hand sides, and
consequently the solutions, are always dense vectors or matrices.

\pk{KSPHPDDM}, the interface between HPDDM Krylov methods and PETSc, is automatically
registered since PETSc version 3.12 when configuring PETSc with the extra
flag \pk{-{}-download-hpddm}. HPDDM linear solvers can be selected using the command
line option \pk{-ksp\_type hpddm}, or the routine \pk{KSPSetType}(ksp,
\pk{KSPHPDDM}), and the following Krylov methods can be accessed:
\begin{itemize}
    \item pseudo-block GMRES or flexible GMRES~\cite{saad1993flexible};
    \item pseudo-block CG or flexible CG~\cite{notay2000flexible};
    \item pseudo-block GCRODR or flexible GCRODR~\cite{carvalho2011flexible};
    \item block GMRES or flexible GMRES~\cite{calandra2012flexible};
    \item block CG~\cite{o1980block};
    \item breakdown-free block CG~\cite{ji2017breakdown};
    \item block GCRODR or flexible GCRODR.
\end{itemize}
While being mathematically equivalent to
their ``standard'' counterparts, the pseudo-block variants try to fuse together multiple similar
operations like matrix--vector
products to achieve higher arithmetic intensity ({\color{red} which method does this?}), or to decrease the
number of global synchronizations needed for scalar products ({\color{red} which method does this?}).
HPDDM Krylov methods may be selected {\color{red} TODO API FOR KSPHPDDMSetType? KSPHPDDMSetVariant?} using the
options \pk{-ksp\_hpddm\_type} \pk{(gmres$|$cg$|$gcrodr$|$bgmres$|$bcg$|$bfbcg$|$bgcrodr$|$preonly)}.
Preconditioning variants can be specified via
\pk{-ksp\_hpddm\_variant} \pk{(left$|$right$|$flexible)} to select whether the preconditioner
has to applied on the left, on the right, or if it cannot be represented as a linear operator.
{\color{red} out of curiosity. how can you choose the side if felixble is specified?}
It is also possible to set the preconditioning side through the more common PETSc option \pk{-ksp\_pc\_side (left$|$right)}.
Note that the value of the latter option is useless for BCG or BFBCG, and HPDDM CG only handles left and
variable preconditioning {\color{red} can you elaborate more?}.
Finally, convergence monitoring with the \pk{KSPMonitor} interface is fully supported, as well as the specification
of customized convergence testing via the \pk{KSPSetConvergenceTest} callback.

According to the standard PETSc notation, ksp($A$, $P_{A'}$)
defines a \pk{KSP} used to solve a system $Ax=b$ using a preconditioner $P_{A'}$ built
using an operator $A'$ which is usually, but not necessarily, equal to $A$. While solving
the linear system via \pk{KSPSolve}(ksp, $b$, $x$) or \pk{KSPMatSolve}(ksp, $B$, $X$),
HPDDM will repeatedly call the following PETSc routines:
\begin{itemize}
    \item \pk{MatMult}($A$, $x$, $y$) for $y = Ax$;
    \item \pk{MatMatMult}($A$, $X$, {\scriptsize \pk{MAT\_REUSE\_MATRIX}}, {\scriptsize
        \pk{PETSC\_DEFAULT}}, $Y$)\footnote{or \pk{MatProductNumeric}($Y$) with PETSc 3.14.0 and above} for $Y =
        AX$;
    \item \pk{PCApply}($P_{A'}$, $y$, $x$) for $x = P_{A'}y$.
\end{itemize}
As of version 3.13.2, PETSc does not have a general routine for applying $P_{A'}$ to a
tall-and-skinny dense matrix $Y$. In HPDDM, a specialized routine has been implemented
to compute $X = P_{A'}Y$ if the preconditioner is of one of the following types:
\begin{itemize}
    \item \pk{PCLU} or \pk{PCCHOLESKY}: exact $LU$ or Cholesky factorization, i.e., $P_{A'} = {A'}^{-1}$;
    \item \pk{PCILU} or \pk{PCICC}: incomplete $LU$ or Cholesky factorization;
    \item \pk{PCBJACOBI}: block Jacobi with exact or incomplete subdomain solvers as above;
    \item \pk{PCASM}: overlapping Schwarz methods with exact or incomplete subdomain solvers as above.
\end{itemize}
In the aforementioned cases, it is indeed possible to access the so-called factored matrix $F$
%via \pk{PCFactorGetMatrix} and then call \pk{MatMatSolve}($F$, $\tilde{Y}$, $\tilde{X}$).
via \pk{PCFactorGetMatrix} and then call \pk{MatMatSolve}($F$, $Y$, $X$) to take advantage
of the blocking solving strategy from the various factorization packages interfaced with PETSc
such as MUMPS~\cite{amestoy2001fully}, SuiteSparse, MKL\_PARDISO and SUPERLU {\color{red} ADD REFERENCES}.
For the types of preconditioners not implementing \pk{PCFactorGetMatrix},
HPDDM will loop over the columns of $Y$ and successively apply $P_{A'}$ to get all columns of $X$.
Future work will consider extending the various implementations of the PETSc
preconditioning infrastructure to increase the arithmetic intensity of the
block preconditioner application phase.

The rest of the operations are then performed directly inside HPDDM {\color{red} what about VecAXPY, VecNorm etc?}.
Specifically, within Krylov methods using the Arnoldi process, or when recycling is requested,
we need $QR$ factorizations of tall-and-skinny dense matrices to orthonormalize
candidate basis vectors. Such operations are performed using the CholQR algorithm~\cite{stathopoulos2002block}, or
via the block variants of the (modified) Gram--Schmidt method.
For Hessenberg matrices generated by the Arnoldi
process, it is common to update their $QR$ decomposition using Givens
rotations~\cite{saad2003iterative}. For block Hessenberg matrices, Householder reflectors are
used instead~\cite{gutknecht2008updating}.
These orthonormalization variants can selected at runtime with the option \pk{-ksp\_hpddm\_qr (cholqr$|$mgs$|$cgs)}. {\color{red} API?}

    \subsubsection{User-defined deflation space} Recycling Krylov methods try to
extract convergence information by solving a small standard or a generalized dense eigenproblem
at the end of each cycle or when convergence is reached and then reuse it appropriately for a
subsequent solve~\cite{saad2000deflated,stathopoulos2002block,soodhalter2014krylov}.
The routine \pk{KSPHPDDMGetDeflationSpace} can be used to inquire about the details of such deflation subspaces, stored as distributed
dense tall-and-skinny matrices. It is also possible to provide this information
via \pk{KSPHPDDMSetDeflationSpace} in case end users already know a good initial
deflation subspace.

@@

{\color{red} TODO: move KSPMatSolve discussion out, and clean up the section}

@@

\bibitem{templateBook}
{\sc R.~Barrett, M.~Berry, T.~F. Chan, J.~Demmel, J.~Donato, J.~Dongarra,
 V.~Eijkhout, R.~Pozo, C.~Romine, and H.~Van der Vorst}
 {\em Templates for the solution of linear systems: building blocks for iterative methods},
  Society for Industrial and Applied Mathematics, (1994).

\bibitem{xsdk}
{\sc R.~Bartlett, I.~Demeshko, T.~Gamblin, G.~Hammond, M.~A. Heroux, J.~Johnson, A.~Klinvex, X.~Li, L.~Curfman~McInnes, J.~D. Moulton, et~al.},
  {\em xSDK Foundations: Toward an Extreme-scale Scientific Software Development Kit},
  Supercomputing Frontiers and Innovations, 4 (2017), pp.~69--82.

@@

\bibitem{Frommer}
{\sc A.~Frommer, K.~Lund, D.~B. Szyld},
 {\em Block Krylov subspace methods for funcions of matrices},
 Electronic Transactions on Numerical Analysis, 47 (2017), pp.~100--126.
